A Method for the Spectroscopic Inference of Stellar Parameters

### 2014-06-14 tex/paper.tex
changed bib tags

%% Nomenclature guide
%%%%%%%%%%%%%%%%%%%%%
% * collections of synthetic spectra are generically called ``libraries.'' The term 
% ``grid'' is used when referring specifically to the spacing of spectral parameters
% in the library

@@

\comm{Important: Note how I have changed the naming convention in the .bib file, so that a paper is 
referenced by the first author's last name and the year, like czekala14.  Your haphazard formatting 
was driving me nuts, and you can't find anything in the bib file (plus, if you read inside the .tex 
file, my approach makes things a hell of a lot easier).}

@@

There are various approaches to synthesizing a spectrum, $f_{\lambda}$, for a specific set of model
parameters, $\vt_{\ast}$.  In an ideal case, a model stellar atmosphere is constructed and then 
subsequently processed through a radiative transfer code \citep[e.g.,][]{kurucz93,hauschildt99}.  
However, in general this approach is still computationally prohibitive for any iterative method of 
probabilistic inference.  One partial compromise is to interpolate over a library of atmosphere 
structures that were pre-computed for a discrete grid of parameter values, $\{\vt_{\ast}\}^{\rm 
grid}$, for some arbitrary $\vt_{\ast}$, and then perform a radiative transfer calculation with 
that interpolated atmosphere to synthesize $f_{\lambda}$ \citep[e.g., as for SME;][]{vp96}.  A more 
common variant is to instead rely on interpolation over a library of pre-synthesized model spectra, 
$f_{\lambda}(\{\vt_{\ast}\}^{\rm grid})$ \citep[e.g.,][]{castelli04,allard12,husser13}.  While 
technically the former approach is most similar to the ideal case, the computational cost of 
repeated spectral synthesis is sufficiently high to make a detailed exploration of parameter space 
(particularly for data with a large spectral range) considerably less appealing.  A related, but 
different approach is to eschew forward modeling entirely (and therefore repeated spectral 
syntheses and/or library interpolations), and instead evaluate the models only at the discrete grid 
points of the library.  Then, these discretized samples of the posterior probability density can be 
interpolated to an arbitrary $\vt_{\ast}$ to construct appropriate confidence intervals 
\citep[e.g., the method of SPC;][]{blj+12}.  The difficulty with this latter approach is that the 
parameter uncertainties can be smaller than the grid spacing; in that case, there is valid concern 
that this interpolation might not accurately recover intrinsic parameter degeneracies.

@@

We adopt a somewhat analogous approach to deal with this issue, but cast it internal to the 
modeling framework to appropriately propagate the uncertainty introduced by additional degrees of 
freedom in the model.  The residual calibration errors are treated as an explicit contribution to 
the model spectrum, enabling us to explore the distribution of possible calibrations (see 
Section~\ref{subsec:MCMC}) and then eventually marginalize out the associated nuisance parameters.  
In essence, the inferences for the relevant stellar parameters will properly account for the 
underlying uncertainty in the calibration process.  This is achieved in practice by distorting the 
model spectrum with a (low-order) Chebyshev polynomial \citep[e.g.,][]{elh+06,koleva09}, 
\begin{equation}
\vM(\vt, \vT_{\rm cheb}) = \vM(\vt) \times \sum_n c_n \, T_n,
\end{equation}
where $T_n$ are the standard Chebyshev functions of order $n$ and the coefficients are treated as a
set of nuisance (hyper-)parameters, $\vT_{\rm cheb} = \{c_n\}^{\rm order}$, for each available 
spectral order in the dataset.  With judicious priors on $\vT_{\rm cheb}$, we can ensure that the 
unintended treatment of real spectral features (e.g., broad, deep molecular bands) as calibration 
artifacts is negligible (see Secton~\ref{sec:examples} for examples).  The lowest-order (scaling) 
coefficient, $c_0$, is by its nature degenerate with the solid angle parameter, $\Omega$.  
Therefore, we enforce an additional constraint by requiring that the mean of the polynomial is 
unity.  For data with a single spectral order, this means setting $c_0 = 1$.  If the goal is to 
model multiple spectral orders, we assign $c_0 = 1$ in an arbitrary order as an anchor, but permit 
the $c_0$ values in other orders to be different (as necessary).  It is worth noting that this 
formalism can, in principle, be extended to develop models for completely uncalibrated spectra, 
rather than the residual calibration mismatches as described here (with a suitable relaxation of 
the priors on $\vT_{\rm cheb}$).  Figure \ref{fig:cheb} offers a practical demonstration of how 
these nuisance parameters are applied. 
% SA: I envision this figure much like the previous one, in a 1-column vertical distribution of 
% panels.  At top, you would show the data spectrum at an overlap of two spectral orders, along 
% with the post-processed model with no Chebyshev polynomials applied.  Next, you would show the 
% two Chebyshev polynomials appropriate for the two orders.  And then in the final panel you would 
% revisit the top panel with the polynomials applied, demonstrating the improvement.

@@

For a well-designed spectrograph, this {\it global} (i.e., present throughout the spectrum) 
covariant structure should have a relatively low amplitude and small correlation length.  To 
describe that structure, we assume a stationary covariance kernel (also called a radial basis 
function) with an amplitude that depends only on the velocity distance between two pixels, 
\begin{equation}
  r_{ij} = \Delta v = \frac{c}{2} \left | \frac{\lambda_i 
   - \lambda_j}{ \lambda_i + \lambda_j} \right |,
\end{equation}
where $c$ is the speed of light.  The kernel is chosen to parametrically describe the covariance 
between pixel residuals, such that
\begin{equation}
  \mathcal{K}_{\rm g}(\lambda_i, \lambda_j) =  \langle \vR_i \; \vR_j \rangle.
  \label{eqn:expectation}
\end{equation}
A variety of kernels have been used in the field of Gaussian processes to parameterize such a 
mildly covariant structure \citep[e.g.,][]{rasmussen05}.  Here we adopt the \matern\ kernel with 
$\nu = 3/2$,
\begin{equation}
  \mathcal{K}_{\rm g}(\lambda_i, \lambda_j \, |\, a_{\rm g},\, \ell) = a_{\rm g} \left(1 + \frac{\sqrt{3}\, r_{ij}}{\ell} \right ) \exp 
   \left (- \frac{\sqrt{3}\, r_{ij}}{\ell} \right ),
\end{equation}
which is parameterized by an amplitude $a_g$ and scale $\ell$ and makes for a smooth transition to 
negligible covariance at large $r$.  To ensure that $\vC$ remains a relatively sparse matrix that 
enables computational expendiency, we employ a Hann window function
\begin{equation}
  w\,(\lambda_i, \lambda_j |\, r_0) = \left \{ 
    \begin{array}{cc}
    \frac{1}{2} + \frac{1}{2} \cos \left(\frac{\pi r_{ij}}{r_0} \right) & r_{ij} \le r_0 \\
    0 & r_{ij} > r_0 \\
  \end{array}
  \right .
  \label{eqn:Hann}
\end{equation}
to taper the kernel (so the effective kernel is $w \, \mathcal{K}_g$).  The truncation distance 
$r_0$ can be fixed to some reasonable multiple of the scale parameter (here we set $r_0 = 4\ell$).  
The examples in Figure \ref{fig:matern} demonstrate that such a kernel readily produces correlated 
structure in the residual spectrum that is similar to those seen in a typical data--model 
comparison.

@@

We use a Markov chain Monte Carlo (MCMC) algorithm coupled to a Gibbs sampler to explore the posterior distribution (Equation~\ref{eqn:lnprob}). The MCMC routine efficiently explores a high dimensional probability space using a stochastic, iterative approach. The Gibbs sampler provides a way to easily sample a large number of parameters in a simple, organized manner by sampling only a subset of parameters at any given time, but then rotating among all of the subsets of parameters. For more on MCMC and Gibbs samplers see \citet[ch 11]{gcs+13} and the references therein.

@@

At each stage of the Gibbs sampler, the likelihood function (Equation~\ref{eqn:lnprob}) is evaluated for a proposal of a particular subset of parameters, conditional on the current values of all the other parameters. This hierarchical parameter structure and the kernel parameterization of the covariance matrix influences how the algorithm may be efficiently implemented in code. For a typical optical spectrum with $\gtrsim$1,000 pixels, the most computationally intensive step of the likelihood evaluation is usually the matrix product $\fR^T C^{-1} \fR$. Since we have designed the covariance matrix to be sparse, we can use optimized sparse matrix algorithms which are much faster and memory efficient than dense matrix operations. Because we are not interested in the matrix inverse $C^{-1}$ by itself, but rather the product with the residual vectors, we can use efficient routines for solving linear systems to bypass the computationally difficult step of matrix inversion. Additionally, because the covariance matrix is positive semi-definite, we can use the Cholesky factorization of the matrix to optimize the evaluation of the matrix product. Once the covariance matrix is factorized, any subsequent evaluation of the matrix product for different residual vectors $\fR$ is extremely rapid. This makes the $\vtstar$ and $\vtcheb$ steps of the Gibbs sampler extremely fast.
When we sample in the nuisance parameters which affect the covariance matrix, we must redo the Cholesky factorization of $C$ for each update. However, because we designed the kernels to deliver a sparse matrix these operations are efficient enough to be used with an MCMC algorithm.  We use the high-performance \texttt{SuiteSparse/CHOLMOD}\footnote{\url{http://www.cise.ufl.edu/research/sparse/cholmod/}} library to implement the sparse matrix and Cholesky factorization operations \citep{cdh+08, dh09} and the Metropolis-Hastings sampler included in the \texttt{Python} MCMC package \texttt{emcee} \citep{fhl+12}.

@@

Linking the covariance matrices of stars could be done serially, where the aggregate covariance matrix of all previous fits is used as a prior for the current iteration. Or, the covariance matrices could instead be linked hierarchically, in that the parameters describing the depth and width of a line residual for a specific star $a_1$ and $\sigma_1$, are modeled as coming from of a population of possible depths and widths for a given spectral type. Each stellar spectrum will have a slightly different realization of a spectral line, which will have some scatter about the average residual height. Linking the covariance matrix between spectra of similar stars allows us to grow more confident in our assessment that certain synthetic spectral lines are indeed outliers and should be appropriately down-weighted. In turn, as we become more certain of the weights, the stellar posteriors will become narrower and make our estimates of the stellar parameters more precise. This ability of the model to mutually inform sets of parameters is one of the major advantages of hierarchical Bayesian analysis \citep{kru10}.

>>>>

@@

@@

There are various approaches to synthesizing a spectrum, $f_{\lambda}$, for a specific set of model
parameters, $\vt_{\ast}$.  In an ideal case, a model stellar atmosphere is constructed and then 
subsequently processed through a radiative transfer code \citep[e.g.,][]{kurucz93,hauschildt99}.  
However, in general this approach is still computationally prohibitive for any iterative method of 
probabilistic inference.  One partial compromise is to interpolate over a library of atmosphere 
structures that were pre-computed for a discrete grid of parameter values, $\{\vt_{\ast}\}^{\rm 
grid}$, for some arbitrary $\vt_{\ast}$, and then perform a radiative transfer calculation with 
that interpolated atmosphere to synthesize $f_{\lambda}$ \citep[e.g., as for SME;][]{valenti96}.  A more 
common variant is to instead rely on interpolation over a library of pre-synthesized model spectra, 
$f_{\lambda}(\{\vt_{\ast}\}^{\rm grid})$ \citep[e.g.,][]{castelli04,allard12,husser13}.  While 
technically the former approach is most similar to the ideal case, the computational cost of 
repeated spectral synthesis is sufficiently high to make a detailed exploration of parameter space 
(particularly for data with a large spectral range) considerably less appealing.  A related, but 
different approach is to eschew forward modeling entirely (and therefore repeated spectral 
syntheses and/or library interpolations), and instead evaluate the models only at the discrete grid 
points of the library.  Then, these discretized samples of the posterior probability density can be 
interpolated to an arbitrary $\vt_{\ast}$ to construct appropriate confidence intervals 
\citep[e.g., the method of SPC;][]{buchhave12}.  The difficulty with this latter approach is that the 
parameter uncertainties can be smaller than the grid spacing; in that case, there is valid concern 
that this interpolation might not accurately recover intrinsic parameter degeneracies.

@@

We adopt a somewhat analogous approach to deal with this issue, but cast it internal to the 
modeling framework to appropriately propagate the uncertainty introduced by additional degrees of 
freedom in the model.  The residual calibration errors are treated as an explicit contribution to 
the model spectrum, enabling us to explore the distribution of possible calibrations (see 
Section~\ref{subsec:MCMC}) and then eventually marginalize out the associated nuisance parameters.  
In essence, the inferences for the relevant stellar parameters will properly account for the 
underlying uncertainty in the calibration process.  This is achieved in practice by distorting the 
model spectrum with a (low-order) Chebyshev polynomial \citep[e.g.,][]{eisenstein+06,koleva09}, 
\begin{equation}
\vM(\vt, \vT_{\rm cheb}) = \vM(\vt) \times \sum_n c_n \, T_n,
\end{equation}
where $T_n$ are the standard Chebyshev functions of order $n$ and the coefficients are treated as a
set of nuisance (hyper-)parameters, $\vT_{\rm cheb} = \{c_n\}^{\rm order}$, for each available 
spectral order in the dataset.  With judicious priors on $\vT_{\rm cheb}$, we can ensure that the 
unintended treatment of real spectral features (e.g., broad, deep molecular bands) as calibration 
artifacts is negligible (see Secton~\ref{sec:examples} for examples).  The lowest-order (scaling) 
coefficient, $c_0$, is by its nature degenerate with the solid angle parameter, $\Omega$.  
Therefore, we enforce an additional constraint by requiring that the mean of the polynomial is 
unity.  For data with a single spectral order, this means setting $c_0 = 1$.  If the goal is to 
model multiple spectral orders, we assign $c_0 = 1$ in an arbitrary order as an anchor, but permit 
the $c_0$ values in other orders to be different (as necessary).  It is worth noting that this 
formalism can, in principle, be extended to develop models for completely uncalibrated spectra, 
rather than the residual calibration mismatches as described here (with a suitable relaxation of 
the priors on $\vT_{\rm cheb}$).  Figure \ref{fig:cheb} offers a practical demonstration of how 
these nuisance parameters are applied. 
% SA: I envision this figure much like the previous one, in a 1-column vertical distribution of 
% panels.  At top, you would show the data spectrum at an overlap of two spectral orders, along 
% with the post-processed model with no Chebyshev polynomials applied.  Next, you would show the 
% two Chebyshev polynomials appropriate for the two orders.  And then in the final panel you would 
% revisit the top panel with the polynomials applied, demonstrating the improvement.

@@

For a well-designed spectrograph and sufficiently accurate model spectrum, this
{\it global} (i.e., present throughout the spectrum) covariant structure should
have a relatively low amplitude and small correlation length.  To describe that
structure, we assume a stationary covariance kernel (also called a radial basis
function) with an amplitude that depends only on the velocity distance between
two pixels, 
\begin{equation}
  r_{ij} = r(\lambda_i, \lambda_j) = \Delta v = \frac{c}{2} \left | \frac{\lambda_i 
   - \lambda_j}{ \lambda_i + \lambda_j} \right |,
\end{equation}
where $c$ is the speed of light.  The kernel is chosen to parametrically describe the covariance 
between pixel residuals, such that
\begin{equation}
  \mathcal{K}_{\rm g}(\lambda_i, \lambda_j) =  \langle \vR_i \; \vR_j \rangle.
  \label{eqn:expectation}
\end{equation}
A variety of kernels have been used in the field of Gaussian
processes to parameterize such a mildly covariant structure
\citep[e.g.,][]{rasmussen05}.  Here we adopt the \matern\ kernel with 
$\nu = 3/2$,
\begin{equation}
  \mathcal{K}_{\rm g}(\lambda_i, \lambda_j \, |\, a_{\rm g},\, \ell) = a_{\rm g} \left(1 + \frac{\sqrt{3}\, r_{ij}}{\ell} \right ) \exp 
   \left (- \frac{\sqrt{3}\, r_{ij}}{\ell} \right ),
\end{equation}
which is parameterized by an amplitude $a_g$ and scale $\ell$ and makes for a smooth transition to 
negligible covariance at large $r$.  To ensure that $\vC$ remains a relatively sparse matrix that 
enables computational expendiency, we employ a Hann window function
\begin{equation}
  w\,(\lambda_i, \lambda_j |\, r_0) = \left \{ 
    \begin{array}{cc}
    \frac{1}{2} + \frac{1}{2} \cos \left(\frac{\pi r_{ij}}{r_0} \right) & r_{ij} \le r_0 \\
    0 & r_{ij} > r_0 \\
  \end{array}
  \right .
  \label{eqn:Hann}
\end{equation}
to taper the kernel (so the effective kernel is $w \, \mathcal{K}_g$).  The truncation distance 
$r_0$ can be fixed to some reasonable multiple of the scale parameter (here we set $r_0 = 4\ell$).  
The examples in Figure \ref{fig:matern} demonstrate that such a kernel readily produces correlated 
structure in the residual spectrum that is similar to those seen in a typical data--model 
comparison.

@@

We use a Markov chain Monte Carlo (MCMC) algorithm coupled to a Gibbs sampler to explore the posterior distribution (Equation~\ref{eqn:lnprob}). The MCMC routine efficiently explores a high dimensional probability space using a stochastic, iterative approach. The Gibbs sampler provides a way to easily sample a large number of parameters in a simple, organized manner by sampling only a subset of parameters at any given time, but then rotating among all of the subsets of parameters. For more on MCMC and Gibbs samplers see \citet[ch 11]{gelman13} and the references therein.

@@

At each stage of the Gibbs sampler, the likelihood function (Equation~\ref{eqn:lnprob}) is evaluated for a proposal of a particular subset of parameters, conditional on the current values of all the other parameters. This hierarchical parameter structure and the kernel parameterization of the covariance matrix influences how the algorithm may be efficiently implemented in code. For a typical optical spectrum with $\gtrsim$1,000 pixels, the most computationally intensive step of the likelihood evaluation is usually the matrix product $\fR^T C^{-1} \fR$. Since we have designed the covariance matrix to be sparse, we can use optimized sparse matrix algorithms which are much faster and memory efficient than dense matrix operations. Because we are not interested in the matrix inverse $C^{-1}$ by itself, but rather the product with the residual vectors, we can use efficient routines for solving linear systems to bypass the computationally difficult step of matrix inversion. Additionally, because the covariance matrix is positive semi-definite, we can use the Cholesky factorization of the matrix to optimize the evaluation of the matrix product. Once the covariance matrix is factorized, any subsequent evaluation of the matrix product for different residual vectors $\fR$ is extremely rapid. This makes the $\vtstar$ and $\vtcheb$ steps of the Gibbs sampler extremely fast.
When we sample in the nuisance parameters which affect the covariance matrix, we must redo the Cholesky factorization of $C$ for each update. However, because we designed the kernels to deliver a sparse matrix these operations are efficient enough to be used with an MCMC algorithm.  We use the high-performance \texttt{SuiteSparse/CHOLMOD}\footnote{\url{http://www.cise.ufl.edu/research/sparse/cholmod/}} library to implement the sparse matrix and Cholesky factorization operations \citep{chen08, davis09} and the Metropolis-Hastings sampler included in the \texttt{Python} MCMC package \texttt{emcee} \citep{foreman-mackey12}.

@@

Linking the covariance matrices of stars could be done serially, where the aggregate covariance matrix of all previous fits is used as a prior for the current iteration. Or, the covariance matrices could instead be linked hierarchically, in that the parameters describing the depth and width of a line residual for a specific star $a_1$ and $\sigma_1$, are modeled as coming from of a population of possible depths and widths for a given spectral type. Each stellar spectrum will have a slightly different realization of a spectral line, which will have some scatter about the average residual height. Linking the covariance matrix between spectra of similar stars allows us to grow more confident in our assessment that certain synthetic spectral lines are indeed outliers and should be appropriately down-weighted. In turn, as we become more certain of the weights, the stellar posteriors will become narrower and make our estimates of the stellar parameters more precise. This ability of the model to mutually inform sets of parameters is one of the major advantages of hierarchical Bayesian analysis \citep{kruschke10}.
